\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-05 k-planar.tex
First major merge done

\newcommand{\PRlabel}[1]{\label{PR:#1}}
\newcommand{\PRref}[1]{(PR\ref{PR:#1})}
\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}

@@

\newcommand{\Bagsize}{\ensuremath{\frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4}}
\newcommand{\bagsize}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 4}}
\newcommand{\treewidth}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 3}}

@@

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $24k^2 + 60k + 36$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

The purpose of this section is to prove \cref{k-planar} and related structural results. 
We begin by introducing a tool for studying planar graphs, called a tripod decomposition, that is implicit in previous work on $H$-partitions of planar graphs \cite{dujmovic.joret.ea:planar}.  We then show how the tripod decomposition of a planarised version of a $k$-planar graph $G$ can be used to find a layered $H$-partition of $G$ of small layered width in which $H$ has small treewidth.

\subsection{Tripod Decompositions}
\label{TripodDecompositions}

Let $T$ be a tree rooted at some node $r\in V(T)$. The \emph{$T$-depth} of a node $x\in V(T)$ is equal to the length of the path from $x$ to $r$ in $T$. A path $P$ in $T$ is \emph{vertical} if, for each integer $i$, $P$ contains at most one vertex of $T$-depth $i$.  The deepest vertex in a vertical path $P$ is called  the \emph{lower endpoint} of $P$ and the shallowest vertex in $P$ is called the \emph{upper endpoint} of $P$. For each node $v\in V(T)$, we call each node on the path, in $T$, from $v$ to $r$ (including $v$ itself) a \emph{$T$-ancestor} of $v$.  If $v'$ is a $T$-ancestor of $v$, then $v$ is a \emph{$T$-descendant} of $v'$.

A \emph{near-triangulation} $N$ is a plane graph whose outer face, $F$, is bounded by a simple cycle in $N$ and each of whose inner faces is bounded by a 3-cycle in $N$.  A \emph{triangulation} is a near-triangulation whose outer face is bounded by a 3-cycle.  For any cycle $F$ in a near-triangulation $N$ we use $N\llparenthesis F\rrparenthesis$ to denote the near-triangulation consisting of all the edges and vertices of $N$ that are contained in $F$ or its interior.

For a connected graph $G$ and two vertices $v,w\in V(G)$, we use $\dist_{G}(v,w)$ to denote the length of a shortest path, in $G$, with endpoints $v$ and $w$.  For a vertex $r\in V(G)$, the \emph{BFS layering} of $G$ from $r$ is the layering $\langle V_0,V_1,\ldots\rangle$ of $G$ where $V_i=\{v\in V(G): \dist_{G}(r,v)=i\}$ for each integer $i\ge 0$.  A \emph{BFS spanning tree} $T$ of $G$ rooted at $r$ is a spanning tree of $G$ rooted at $r$ where, for every $v\in V(G)$, $\dist_T(v,r)=\dist_G(v,r)$.

For the remainder of this section, $\Delta$ is a triangulation, $r$ is a degree-3 vertex on the outer face of $\Delta$, and $T$ is a BFS spanning tree of $\Delta$ rooted at $r$.

Fix some cycle $F$ in $\Delta$ and let $N=\Delta\llparenthesis F\rrparenthesis$.
Let $\tau=v_1v_2v_3$ be an inner face of $N$ and consider the three minimal vertical paths $Q_1$, $Q_2$, and $Q_3$ in $T$ such that, for each $i\in\{1,2,3\}$ the lower endpoint of $Q_i$ is $v_i$ and the upper endpoint $v_i'$ of $Q_i$ is in $V(F)$.  If $Q_1$, $Q_2$, and $Q_3$ are vertex-disjoint, then the graph $\overline{Y}:=\tau\cup Q_1\cup Q_2\cup Q_3$ is called the \emph{closed tripod} in $N$ determined by $\tau$.  We call $Q_1$, $Q_2$, and $Q_3$ the \emph{legs}, $v_1'$, $v_2'$ and $v_3'$ the \emph{feet}, and $\tau$ the \emph{crotch} of $\overline{Y}$.  We call $Y:=\overline{Y}-F$ the \emph{(open) tripod} of $N$ determined by $\tau$.  The graph $M:=\overline{Y}\cup F$ is called a \emph{Mercedes graph} and the (at most three) inner faces of $M$ other than $\tau$ are called the \emph{wedges} of $M$.

Let $r$ be a degree-3 node on the outer face of a triangulation $\Delta$.  A \emph{tripod decomposition}  $K=K(\Delta, T, r)$ is a rooted tree in which each non-root node $x\in V(K)$ is associated with several objects (refer to \cref{K-node}):

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \includegraphics{figs/K-node-1} &
      \includegraphics{figs/K-node-2}
    \end{tabular}
  \end{center}
  \caption{The elements associated with a node $x\in V(K)$.}
  \label{K-node}
\end{figure}

\begin{enumerate}
  \item A cycle $F_x$ in $\Delta$ and the near-triangulation $N_x:=\Delta\llparenthesis F_x\rrparenthesis$.

  \item An inner face $\tau_x$ of $N_x$ that determines a closed tripod $\overline{Y}_x$ and an open tripod $Y_x:=\overline{Y}_x-F_x$ in $N_x$, both with crotch $\tau_x$.

  \item The \emph{Mercedes graph} $M_x:=F_x\cup \overline{Y}_x$
\end{enumerate}

The root node $r^0_K$ of $K$ is exceptional.  We define $F_{r^0_K}$ to be the single-vertex graph containing $r$, we define $\overline{Y}_{r^0_K}$ to be the star formed by $r$ and its three neighbours in $\Delta$, and (as usual) $Y_{r^0_K}:=\overline{Y}_{r^0_K}-F_{r^0_K}$.  The root node $r^0_K$ has exactly one child $r_K$, called the \emph{sub-root} of $K$, for which $F_{r_K}$ is the 3-cycle formed by the neighbours of $r$ in $\Delta$.

The non-root nodes of the tripod decomposition $K$ (including the sub-root $r_K$) satisfy the following conditions:
\begin{enumerate}[(PR1)]
  \item \PRlabel{children} For each non-root node $x\in V(K)$ such that $M_x$ has $f\le 3$ wedges $F_{x,1},\ldots,F_{x,f}$, the node $x$ has exactly $f$ children $y_1,\ldots,y_f$ in $K$ and $F_{y_i} = F_{x,i}$ for each $i\in\{1,\ldots,f\}$.

  \item \PRlabel{ancestor-boundary} For each non-root node $x\in V(K)$, there exists a partition of $V(F_x)$ into $g\le 3$ non-empty sets $P_1,\ldots,P_g$ such that, for each $i\in\{1,\ldots,g\}$, $F[P_i]$ is a path that is contained in $Y_a$ for some $K$-ancestor $a$ of $x$.

  \item \PRlabel{tripod-feet} For each $i\in\{1,\ldots,g\}$, at least one vertex in $P_i$ is a foot of the closed tripod $\overline{Y}_x$.

  \item \PRlabel{partition} $\{Y_x : x\in V(K)\}$ is a partition of $V(\Delta)\setminus\{r\}$.
\end{enumerate}

For any triangulation $\Delta$ in which the outer face has a vertex $r$ of degree 3 and for any BFS spanning tree of $\Delta$ rooted at $r$, there exists a tripod decomposition $K=K(\Delta,T,r)$.  Indeed, this tree is constructed (albeit implicitly) by \citet[Proof of Lemma~14]{dujmovic.joret.ea:planar}.\footnote{\citet{dujmovic.joret.ea:planar} do not require that $T$ be a BFS spanning tree, only that its root, $r$ has degree 3 in $\Delta$ and in $T$.}

% The following claim is not used until later in the paper. However, it's proof is a helpful exercise in understanding the properties of a PS-Tree.
% 
% \begin{clm}\label{strong-ancestor-boundary}
%   Let $K$ be a PS-tree and
%   let $y\in V(K)$ be a non-root node with parent $p$ such that $V(F_y\cap Y_a)\neq\emptyset$ for some $K$-ancestor $a\neq p$ of $p$. Then $V(F_p\cap Y_a)\neq\emptyset$.
% \end{clm}
% 
% \begin{proof}
%   By \propref{children}, $F_y$ is a wedge in the Mercedes graph $M_p$, so $V(F_y)$ can be partitioned into  $P_1'=V(F_y\cap Y_p)$ and $P_{2,3}'=V(F_y\cap F_p)$.  Let $P_1,\ldots,P_g$ denote the partition of $V(F_p)$ given by \propref{ancestor-boundary}.  
% 
%   We claim that \PRref{tripod-feet} ensures that $P_{2,3}'\subset P_i\cup P_j$ for some $i,j\in \{1,\ldots,g\}$.  If $g=2$, this claim is immediate, so assume $g=3$. Then, $F_y\cap \overline{Y}_p$ is the path formed by two legs $Q_2$ and $Q_3$ of $\overline{Y}_p$ and one edge $v_2v_3$ of $\tau_p$.  
%   By \PRref{tripod-feet} and the assumption that $q=3$, $Q_2$ has a foot $v_2'\in P_i$ and $Q_3$ has a foot $v_3'\in P_j$, for some $i\neq j$.  The third foot $v_1'$ of $\overline{Y}_p$ is therefore in $P_\ell$ for the unique $\ell \in\{1,2,3\}\setminus \{i,j\}$.  It follows that $V(P_\ell\cap F_y)=\emptyset$.
% 
%   Setting $P_2'=V(P_i\cup F_y)$ and $P_3'=V(P_j\cup F_y)$ yields a partition $\{P_1',P_2',P_3'\}$ of $V(F_y)$ such that $P_1'\subset V(Y_{p})$,   
%   $P_2'\subset V(Y_{a_2})$, and $P_3'\subset V(Y_{a_3})$  for some $K$-ancestors $a_2$ and $a_3$ of $a$.  Therefore $a_2$ and $a_3$ are the only $K$-ancestors $a\neq p$ of $p$, such that $V(F_y\cap Y_a)\neq\emptyset$.
%   Now, each foot $v_2',v_3'\in V(F_p)$.  Furthermore, $v_2'\in P_i\subseteq V(Y_{a_2})$ and $v_3'\in P_j\subseteq V(Y_{a_3})$. Therefore, $V(F_p\cap Y_{a_2})\supseteq\{v_2'\}\neq\emptyset$ and $V(F_p\cap Y_{a_3})\supseteq\{v_3'\}\neq\emptyset$, as claimed.
% \end{proof}
% 
% Contained in the proof of \clmref{strong-ancestor-boundary} is the following addendum to \propref{ancestor-boundary}:
% 
% \begin{enumerate}[(PR1')]\setcounter{enumi}{1}
%   \item \proplabel{ancestor-boundary-ii} Furthermore, if $g=3$, then  $P_i=V(F_x\cap Y_p)$ where $p$ is the parent of $x$ in $K$, for one $i\in\{1,2,3\}$.
% \end{enumerate}

% \subsubsection{Layered $H$-Partitions from PS-Trees}
% \seclabel{k-to-h}
% 
% Before continuing, it is worth pointing out how an $H$-partition of $\Delta-\{r\}$ can be obtained from a PS-tree $K=K(\Delta,T,r)$. (Note: There is nothing new in this section.  This is just an alternative presentation of material from \citet{dujmovic.joret.ea:planar}.) Given $K=K(\Delta,T,r)$, we obtain a layered $H$-partition $(\mathcal{L},\mathcal{P})$ as follows. 
% 
% We use the layering $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ where $V_i=\{ v\in V(\Delta)\setminus\{r\}: \dist_{\Delta}(r,v)=i\}$ and $h=\max\{\dist_\Delta(r,v):v\in V(\Delta)\}$.  The fact that $T$ is a BFS tree ensures that $\mathcal{L}$ is, indeed, a layering of $\Delta-\{r\}$.  We use the vertex partition $\mathcal{P}=\{ S_x:=V(Y_x) : x\in V(K)\}$.  \PRref{partition} ensures that $\mathcal{P}$ is, indeed a partition of $V(\Delta)-\{r\}$.
% 
% The resulting layered $H$-partition $(\mathcal{L},\mathcal{P})$ has layered width at most 3 because, for each $x\in V(K)$, $S_x$ is made up of vertices from at most 3 vertical paths in $T$.  These are either the three vertices of $Y_{r_k}$ (if $x=r_K$) or the vertices of the open tripod $Y_x$ whose vertices are included in three vertical paths $Q_1,Q_2,Q_3$ that form the legs of the closed tripod $\overline{Y}_x$.  Therefore $|S_x\cap V_i|\le 3$.
% 
% To show that the quotient graph $H=\Delta/\mathcal{P}$ has small treewidth, we produce a tree-decomposition $\mathcal{K}=\{B_x: x\in V(K)\}$ of $H$ using the tree $K$.  For each $x\in V(K)$, the subtree $K[x]=K[\{y:x\in B_y\}]$ contains
% $x$ as well as every node $y$ such that $V(Y_x)\cap V(F_y) \neq\emptyset$.
% It is straightforward to verify that $K[x]$ is connected for each $x\in V(K)$. The fact that $F_x$ is a cycle in $\Delta$ and \propref{ancestor-boundary} ensures that, if $xy\in E(H)$, then one of $x$ or $y$ is a $K$-ancestor of the other. \propref{ancestor-boundary} also implies that, for any node $y\in V(K)$, $|B_y|\le 4$, since $B_y$ contains $y$ and at most 3 $K$-ancestors of $y$.
% 
% Therefore $(\mathcal{L},\mathcal{P})$ is a layered $H$-partition of $\Delta-\{r\}$ with layered width 3 in which $H$ has treewidth 3.  Furthermore, for each $P\in\mathcal{P}$, $\Delta[P]$ is connected.  Since $\Delta$ is planar, this implies that $H$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}

>>>>

\newcommand{\PRlabel}[1]{\label{PR:#1}}
\newcommand{\PRref}[1]{(PR\ref{PR:#1})}

\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}

\newcommand{\tlabel}[1]{\label{t:#1}}
\newcommand{\tref}[1]{(T\ref{t:#1})}
\newcommand{\ylabel}[1]{\label{y:#1}}
\newcommand{\yref}[1]{(Y\ref{y:#1})}

@@

\newcommand{\Bagsize}{\ensuremath{\frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4}}
% \newcommand{\bagsize}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 4}}
\newcommand{\treewidth}{\ensuremath{\binom{k+4}{3}}-1}

@@

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $15k^2 + 30k$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

The purpose of this section is to prove \cref{k-planar} and related structural results.  We begin with a more general abstraction called $(k,d)$-shortcut systems and prove a structural result, \cref{modern-major-general}, for graphs derived from $(k,d)$-shortcut systems.  The results on $k$-planar (\cref{k-planar}) and $(g,k)$-planar graphs (\cref{g-k-planar}) can then be obtained from \cref{modern-major-general} and existing results for planar and bounded-genus graphs.

\subsection{Shortcut Systems}

For a rooted tree $T$ and any node $x\in V(T)$, we define $T_x := T[\{y\in V(T):\mbox{$x$ is a $T$-ancestor of $y$}\}]$ to be the maximal subtree of $T$ rooted at $x$.  We begin with a fairly standard lemma about normalized tree decompositions:

\begin{lem}\lemlabel{nice-decomposition}
  For any graph $H$ of treewidth $t$, there is a rooted tree $T$ with $V(T)=V(H)$ and a $T$-decomposition $(B_x:x\in V(T))$ of $H$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    % \item\tlabel{rooted}\tlabel{first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    % \item\tlabel{node-set} $V(T)= V(H)$.
    % \item\tlabel{diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y|\le 1$.
    \item\tlabel{subtree-root} For each $x\in V(H)$, the subtree $T[x]:=T[\{y\in V(T):x\in B_y\}]$ is rooted at $x$.
    \item\tlabel{ancestor-edge}\tlabel{last} For each edge $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with any width-$t$ tree decomposition $(B_x:x\in V(T_0))$ of $H$ that uses some rooted tree $T_0$.  Select any node $x\in V(T_0)$, add a leaf $x_0$, with $B_{x_0}=\emptyset$, adjacent to $x$ and root $T_0$ at $x_0$.  Let $f:V(H)\to V(T)$ be function that maps each $x\in V(H)$ onto the root of the subtree $T_0[x]:=T_0[\{y\in V(T_0): x\in B_y]$.  If $f$ is not one-to-one, then select some distinct pair $x,y\in V(H)$ with $a:=f(x)=f(y)$.  Subdivide the edge between $a$ and its parent in $T$ by introducing a new node $a'$ with $B_{a'}=B_{a}\setminus\{x\}$.  This modification reduces the number of distinct pairs $x,y\in V(H)$ with $f(x)=f(y)$, so repeatedly performing this modification will eventually produce a tree-decomposition $(B_x:x\in V(T_0))$ of $H$ in which $f$ is one-to-one.

  Next, consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.

  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.  To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that, if $xy\in E(H)$, then at least one of $x$ or $y$ is contained in $B_z$ for every node $z$ on the path from $x$ to $y$ in $T$.  If neither $x$ nor $y$ is an ancestor of the other, then some node $z$ on this path has $T$-depth less than that of $x$ and $y$.  If $x\in B_z$ this contradicts the fact that $x$ is the root of $T[x]$.  If $y\in B_z$ this contradicts the fact that $y$ is the root of $T[y]$.
\end{proof}

\subsubsection{Generalized Tripod Partitions}

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret these objects as a hierarchical decomposition of $G$.

\begin{clm}\clmlabel{generalized-tripod}
  For each $x\in V(T)$, let $V_x = \bigcup_{y\in V(T_x)} Y_y$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$.  Then, 
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $\ell$.
    \item\ylabel{separator} For every $x\in V(T)$, there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} For every $x\in V(T)$, $x$ has $t'\le t$ strict $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{clm}

Before proving \clmref{generalized-tripod} we point out more properties that are immediately implied by it:

\begin{compactenum}[(Y1)]\setcounter{enumi}{3}
  \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$. 
  \item\ylabel{containment-i} $V_x\subseteq V_a$ for every $T$-ancestor $a$ of $x$. 
  \item\ylabel{containment-ii}$N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
\end{compactenum}

Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices in $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).

\begin{proof}[Proof of \clmref{generalized-tripod}]
  Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$ and the fact that $(\mathcal{L},\mathcal{Y})$ is a layered partition of layered width $\ell$.  Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.

  To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}=Q_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}=Q_a$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path, in $T$ from $z_0:=x'$ to $z_r:=a$.  For each $i\in\{0,\ldots,r\}$, at least one of $a$ or $x'$ is in $B_{z_i}$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{x_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{x_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
  Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $C_x$ contains $x$.
\end{proof}

\subsubsection{Shortcut Systems}

% For non-negative integers $k$ and $d$, a graph $G$ and distinct vertices $v,w\in V(G)$, a \emph{$(k,d)$ $vw$-shortcut} in $G$ is a path in $G$ with endpoints $v$ and $w$ of length at most $k$ and whose internal vertices each have degree at most $d$.  The \emph{$(k,d)^\text{th}$ power of $G$}, denoted $G^{k,d}$, is the graph with vertex $V(G^{k,d})=V(G)$ in which, for each distinct pair $v,w\in V(G)$, $vw\in E(G^{k,d})$ if and only if $G$ contains a $(k,d)$ $vw$-shortcut.

A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if every path in $\mathcal{P}$ has length at most $k$ and if, for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$. Each path in $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, we use $G^{\mathcal{P}}$ to denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

In this section we will show how, given a $(k,d)$-shortcut system $\mathcal{P}$, a layered $H$-partition of $G$ can be used to obtain a layered $I$-partition of $G^{\mathcal{P}}$ where the layered width of the partition does not increase dramatically and the treewidth of $I$ is not much more than the treewidth of $H$.

For convenience, it will be helpful to assume that $\mathcal{P}$ contains a
length-1 $vw$-shortcut for every edge $vw\in E(G)$.  This assumption has no effect on the resulting graph $G^{\mathcal{P}}$ but avoids special cases in some of our proofs.

As in the rest of this section, $(\mathcal{L},\mathcal{Y})$ denotes a layered $H$-partition of $G$; $T$ is a tree used in a tree-decomposition $(B_x:x\in V(T))$ of $H:=G/\mathcal{Y}$ satisfying the conditions of \lemref{nice-decomposition}; and, for each $x\in V(T)$, $Y_x$, $V_x$, $F_x$, and $N_x$ are defined as in \clmref{generalized-tripod}.

For a node $x\in V(T)$, we say that a shortcut $P\in\mathcal{P}$ \emph{crosses} $x$ if $Y_x$ contains an internal vertex of $P$, that is, $P=(v_0,\ldots,v_r)$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \emph{participates} in $x$ if $v\in Y_x$ or $\mathcal{P}$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$. We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

\begin{clm}
  For any $v\in V(G)$ there exists a (unique) node $a(v)\in X_v$ such that
  $a(v)$ is a $T$-ancestor of every node in $X_v$.
\end{clm}

\begin{proof}
  Let 
  \[
     Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:\mbox{$(v_0,\ldots,v_r)\in\mathcal{P}$, $v\in \{v_0,\ldots,v_r\}$}\}
  \]
  Clearly $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.

  We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\mathcal{P}$ that crosses $x$.

  Let $Z_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  
  %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other. 
  Choose $a(v)\in X_H$ to be any member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that $a(v)$ is a $T$-ancestor of every node $x\in X_H$, as required.
\end{proof}

For each $x\in V(T)$, we define the \emph{separator}
\[  
     S_x := \{v\in V(G): a(v)= x\} \enspace .
\]
Observe that $\mathcal{S}:=(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $I:=G^\mathcal{P}/\mathcal{S}$ denote the resulting quotient graph and we let $V(I)\subseteq V(T)$ in the obvious way, so that each $x\in V(I)$ is the vertex obtained by contracting $S_x$ in $G^{\mathcal{P}}$. (Nodes $x\in V(T)$ with $S_x=\emptyset$ do not contribute a vertex to $I$.)

\begin{clm}\clmlabel{s-subset}
  For every $x\in V(T)$, $S_x\subseteq V_x$.
\end{clm}

\begin{proof}
  For the sake of contradiction, assume otherwise, so there exists some $v\in S_x\setminus V_x$.

  By \yref{y-subsets}, $Y_x\subseteq V_x$, so $v\not\in Y_x$.  Therefore, $\mathcal{P}$ contains a path $P$ that crosses $x$.  The path $P$ contains a subpath $v_0,v_1,\ldots,v_{r}$ such that $v=v_0$ and $v_r\in Y_x$. Since $v\not\in V_x$ and $v_r\in Y_x\subseteq V_x$, \yref{separator}, implies that $v_i\in F_x$ for some $i\in\{0,\ldots,r-1\}$. Now \yref{ancestor-edge} implies $v_i\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  Therefore, either $v\in Y_a$ or $P$ crosses $a$. But this implies that $a(v)$ is an ancestor of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

Next we show that $\mathcal{S}$ has small layered width with respect to $\mathcal{L}$:

\begin{clm}\clmlabel{general-width}
  For each $i\in\{1,\ldots,h\}$ and each $x\in V(I)$, $|S_x\cap V_i|\le d\ell(k^2+k+4)$.  
\end{clm}

\begin{proof}
  We will count the number of vertices in $S_x\cap V_i$ by upper-bounding the number of vertices contributed to $S_x\cap V_i$ by each vertex $w\in Y_x$.
  If $w\in Y_x\cap V_i$ and no path in $\mathcal{P}$ includes $w$ as an internal vertex then $w$ contributes only one vertex, itself, to $S_x\cap V_i$.

  Otherwise, consider some path $P\in\mathcal{P}$ that contains $w$ as an internal vertex.  If $w\in V_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap V_i$.  If $w\in V_{i-1}\cup V_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap V_i$. If $w\in V_{i-j}\cup V_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap V_i$.

  For any $j$, the number of vertices $w\in V_{i+j}\cap Y_x$ is at most $\ell$. Each such vertex $w$ is an internal vertex of at most $d$ paths in $\mathcal{P}$. Therefore, 
  \[  |S_x\cap V_i|\le d\ell\cdot\left(k+1 + k + \sum_{j=2}^k 2(k-j)\right) %= d(2k+1) + \sum_{i=1}^{k-2} i
      = d\ell(k^2 +k + 4) \enspace . \qedhere
  \]
\end{proof}

\begin{clm}\clmlabel{i-ancestor}
  For each edge $xy\in E(I)$, one of $x$ or $y$ is a $T$-ancestor of the other.
\end{clm}

\begin{proof}
  Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(I)$, $G^\mathcal{P}$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $xy\in E(G^{\mathcal{P}})$,  $\mathcal{P}$ contains a $vw$-shortcut $P$.  By \clmref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge} $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

\begin{clm}\clmlabel{general-bag-size}
  The graph $I$ has a tree-decomposition in which every bag has size at most $\binom{k+t}{t}$.
\end{clm}

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $I$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $I$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \clmref{i-ancestor} ensures that, for every edge $ax'\in E(I)$, $a,x'\in C_{x'}$ and the connectivity of $T[a]:=T[\{x\in V(T):a\in C_x]$ is obvious.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $I$.  What remains is to bound the size, $\max\{|C_x|:x\in V(T)\}$, of the largest bag.

  Consider an arbitrary node $x\in V(T)$ whose ancestors are $x_0,\ldots,x_{r-1},x_r$ where $x_0$ is the root of $T$ and $x_r:=x$.  To avoid triple-subscripts in what follows, we abuse notation slightly by using $V_i$, $F_i$, and $N_i$,  as shorthands for $V_{x_i}$, $F_{x_i}$ and $N_{x_i}$, respectively.
